When do the effects of interaction happen? Interacting and disturbed galaxies in the COSMOS field

### 2023-12-11 paper-source/main.tex
Updated .tex source with new figure and integrated to the paper.

Once matched, we remove any sources with unphysical photometric measurements for the stellar mass or star formation rates. We then further reduce our sample by only keeping sources within the a mass range of $6.5 \leq \log_{10} \text{M}_{*}(\text{M}_{\odot}) \leq 12.5$ and a star formation rate range of $-5 < \log_{10} \text{SFR} (M_{\odot}yr^{-1}) \leq 3.5$. To also nullify the effect of surface brightness dimming that may occur when attempting to classify tidal features of interacting galaxies, we institute a redshift cut of $z \leq 1.2$. This also makes our sample match the catalogue that will be later used to investigate the incidence of our sample of interacting galaxies with environment (detailed in Section \ref{data:environ}). Applying these cuts reduces our sample size to 3,689 interacting galaxies.

@@

To classify a secondary galaxy for our primary one, it had to be within the cutout we were visually assessing and within the recorded error of the primary photometric redshift. Using photometric redshift cutoffs in this way is often done when calculating environment parameters \citep[e.g][]{2006MNRAS.373..469B} or defining interacting galaxies by close pairs \citep[e.g][]{2022ApJ...940....4S}. A null redshift is defined as one outwith our redshift limits or NaN. A minority of the cutouts we visually assessed were found to have the incorrect primary at the centre. In these cases, we record the correct priamry galaxy ID and extract the ancillary data from the COSMOS2020 catalogue. We then attempt to find the secondary galaxy again for the corrected primary.

@@

\subsection{Classifying Stage of Interaction}\label{sec:staging}
The primary goal of this work is to find if a relation exists between a host of galactic parameters, underlying physical processes and the stage of the interaction. Each stage covers a different part of the dynamical history in an interaction and, in this case, we define it based on the morphology and projected separation between systems. WE define four dinstinct stages of interaction:

@@

\noindent Figure \ref{fig:stages} shows the original OR23 source cutouts of an example galaxy in each stage. There are degeneracies associated with this staging system, however. In this context, we define a degeneracy as when the interacting galaxies may be at two or more parts of the dynamical timescale and we have no way to tell without further information on the system.

@@

Our four stage approach is not a new one, and many other works have utilised it to differentiate different parts of the dynamical history of a galaxy interaction \citep{2022ApJ...937...97C, 2023ApJ...952..122G}. We have noted the degeneracy of each stage as we have described them. Thus, we can now put this fully into the context of the dynamical timescale. Over a typical interaction, we would expect the galaxy pair to always move from stage 1 to stage 2 in the early times of the dynamical timescale. Then, dependent on the velocity of the system, the galaxy pair will either move straight into stage 4 of classification and begin to merge or it will move into stage 3. This change from stage 2 into stage 3 can then take many branching paths dependent on the elocity in the system. If the galaxies have sufficient velocity, stage 3 will be their end state until the tidal features slowly dissipate. If they do not have velocity to escape on another, the system will move from stage 3 back into stage 2. This could then happen for many cycles until finally the galaxies enter stage 4 and coalesce. Figure \ref{fig:illustration} shows the branching paths that the galaxies can take through each stage. Note, these are created using the PySPAM restricted numerical simulation (O'Ryan et al. in prep.) and are for illustrative purposes only.

\begin{figure*}
    \centering
    \includegraphics{}
    \caption{Sorry, I wan't able to create this figure in the editing phase of this paper! But this will be added as soon as I have a small breather, absolutely love the idea of it.}
    \label{fig:illustration}
\end{figure*}

@@

\begin{figure}
    \centering
    \includegraphics[width = 0.45\textwidth]{figures/sfr-mass-stages.pdf}
    \caption{The LePhare stellar mass against the EAzY SFR across the different stages of the interaction. The blue contours are 8 levels of density of the underlying populations in each frame. \textit{Top left}: The stellar mass and star formation rate of Stage 1 of our sample. Here, the interacting galxies are simply close pairs with little to no morphological disturbance. There are clearly two populations here: a main, star forming sequence forming the main population and a smaller red sequence. \textit{Top right}: Stage 2 of the interaction, where the two interactors are close to pericentre. The star forming sequence remains, but the red sequence is reduced significantly. \textit{Bottom left}: Stage 3 of the interaction, where the interactors are close to apocentre or escaped. Here, we see the almost complete disappearance of the red sequence. \textit{Bottom Right}: Stage 4 of the interaction, where the two systems are close to coalescence or have merged. The red sequence of galaxies has completely disappeared.}
    \label{fig:sfr-mass}
\end{figure}

@@

Such a conclusion would also, therefore, explain the often quite large divide in the literature between whether interaction actually leads to enhancement or not. The only part of the interaction which actually causes the enhancement, and is the forking point, is that of stage 2 - where the two galaxies overlap and are at pericentre. From this point, if the galaxies then move off to stage 3 and escape, we will see a gradual decline in the SFE with the stage 3 galaxy only minorly increased in its star formation classification. Whereas, if the galaxies move into a stage 4 interaction and merge, that is where we see the catastrophic results of a major starburst and the complete using and of all the gas in the systems. Thus, leading to a large fraction of quenched post-merger galaxies, but only after the initial coalescence.

@@

However, to more fully study this, we would need a larger sample of confirmed AGN and star forming galaxies from existing photometry or catalogues to make more decisive conclusions based on stage. In this work, we were limited to a small number of AGN and star forming galaxies across each stage. When compared to our paired sample, we do not have the numbers to also divide into stage again. Therefore, Figure \ref{fig:sfg-agn-proj} shows the global AGN number count with projected separtion. We show that the two peaks are due to different stages of the interaction, however, these are using very low number counts and should be treated carefully.

>>>>

Once matched, we remove any sources with non-physical photometric measurements for the stellar mass or star formation rates. We then further reduce our sample by only keeping sources within the a mass range of $6.5 \leq \log_{10} \text{M}_{*}(\text{M}_{\odot}) \leq 12.5$ and a star formation rate range of $-5 < \log_{10} \text{SFR} (M_{\odot}yr^{-1}) \leq 3.5$. To also nullify the effect of surface brightness dimming that may occur when attempting to classify tidal features of interacting galaxies, we institute a redshift cut of $z \leq 1.2$. This also makes our sample match the catalogue that will be later used to investigate the incidence of our sample of interacting galaxies with environment (detailed in Section \ref{data:environ}). Applying these cuts reduces our sample size to 3,689 interacting galaxies.

@@

To classify a secondary galaxy for our primary one, it had to be within the cutout we were visually assessing and within the recorded error of the primary photometric redshift. Using photometric redshift cutoffs in this way is often done when calculating environment parameters \citep[e.g][]{2006MNRAS.373..469B} or defining interacting galaxies by close pairs \citep[e.g][]{2022ApJ...940....4S}. A null redshift is defined as one outwith our redshift limits or NaN. A minority of the cutouts we visually assessed were found to have the incorrect primary at the centre. In these cases, we record the correct primary galaxy ID and extract the ancillary data from the COSMOS2020 catalogue. We then attempt to find the secondary galaxy again for the corrected primary.

@@

\subsection{Classifying Stage of Interaction}\label{sec:staging}
The primary goal of this work is to find if a relation exists between a host of galactic parameters, underlying physical processes and the stage of the interaction. Each stage covers a different part of the dynamical history in an interaction and, in this case, we define it based on the morphology and projected separation between systems. We define four distinct stages of interaction:

@@

\noindent Figure \ref{fig:stages} shows the original OR23 source cutouts of an example galaxy in each stage. There are degeneracy's associated with this staging system, however. In this context, we define a degeneracy as when the interacting galaxies may be at two or more parts of the dynamical timescale and we have no way to tell without further information on the system.

@@

Our four stage approach is not a new one, and many other works have utilised it to differentiate different parts of the dynamical history of a galaxy interaction \citep{2022ApJ...937...97C, 2023ApJ...952..122G}. We have noted the degeneracy of each stage as we have described them. Thus, we can now put this fully into the context of the dynamical timescale. Over a typical interaction, we would expect the galaxy pair to always move from stage 1 to stage 2 in the early times of the dynamical timescale. Then, dependent on the velocity of the system, the galaxy pair will either move straight into stage 4 of classification and begin to merge or it will move into stage 3. This change from stage 2 into stage 3 can then take many branching paths dependent on the velocity in the system. If the galaxies have sufficient velocity, stage 3 will be their end state until the tidal features slowly dissipate. If they do not have velocity to escape on another, the system will move from stage 3 back into stage 2. This could then happen for many cycles until finally the galaxies enter stage 4 and coalesce. Figure \ref{fig:illustration} shows the branching paths that the galaxies can take through each stage. Note, these are created using the Advanced Python Stellar Animation Module restricted numerical simulation \citep[][O'Ryan et al., in prep.]{2016A&C....16...26W} and are for illustrative purposes only.

\begin{figure*}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/stage-evolution.jpg}
    \caption{The progression through an interaction using our stage definitions. In stage 1, we have two systems that are approaching each other but exhibit no tidal features. This is before the point of closest approach has occurred. This is followed by an intial stage 2: the systems are approximately at their closest approach. This stage is often mistaken, in the COSMOS catalogue, for being of only one source. Clear tidal features exist with major disturbance in the two disks. This is followed by stage 3, where there are two distinct cores with clear tidal features. However, after this point, there are two outcomes to the system depending on the galactic velocities. If the secondary has the escape velocity, the system will remain a stage 3 interaction until the tidal features dissipate (and no longer are in our sample). If they do not have enough energy, the system will return to stage 2 of the encounter and then begin to merge at stage 4. Images are from the Advanced Python Stellar Particle Annimation Module interacting galaxy algorithm described in O'Ryan et al. (in prep.) and based on the stellar particle animation module algorithm described in \citet{2016A&C....16...26W}.}
    \label{fig:illustration}
\end{figure*}

@@

\begin{figure}
    \centering
    \includegraphics[width = 0.45\textwidth]{figures/sfr-mass-stages.pdf}
    \caption{The LePhare stellar mass against the EAzY SFR across the different stages of the interaction. The blue contours are 8 levels of density of the underlying populations in each frame. \textit{Top left}: The stellar mass and star formation rate of Stage 1 of our sample. Here, the interacting galaxies are simply close pairs with little to no morphological disturbance. There are clearly two populations here: a main, star forming sequence forming the main population and a smaller red sequence. \textit{Top right}: Stage 2 of the interaction, where the two interactors are close to pericentre. The star forming sequence remains, but the red sequence is reduced significantly. \textit{Bottom left}: Stage 3 of the interaction, where the interactors are close to apocentre or escaped. Here, we see the almost complete disappearance of the red sequence. \textit{Bottom Right}: Stage 4 of the interaction, where the two systems are close to coalescence or have merged. The red sequence of galaxies has completely disappeared.}
    \label{fig:sfr-mass}
\end{figure}

@@

Such a conclusion would also, therefore, explain the often quite large divide in the literature between whether interaction actually leads to enhancement or not. The only part of the interaction which actually causes the enhancement, and is the forking point, is that of stage 2 - where the two galaxies overlap and are at pericentre. From this point, if the galaxies then move off to stage 3 and escape, we will see a gradual decline in the SFE with the stage 3 galaxy only having a minor increase in its star formation classification. Whereas, if the galaxies move into a stage 4 interaction and merge, that is where we see the catastrophic results of a major starburst and the complete using and of all the gas in the systems. Thus, leading to a large fraction of quenched post-merger galaxies, but only after the initial coalescence.

@@

However, to more fully study this, we would need a larger sample of confirmed AGN and star forming galaxies from existing photometry or catalogues to make more decisive conclusions based on stage. In this work, we were limited to a small number of AGN and star forming galaxies across each stage. When compared to our paired sample, we do not have the numbers to also divide into stage again. Therefore, Figure \ref{fig:sfg-agn-proj} shows the global AGN number count with projected separation. We show that the two peaks are due to different stages of the interaction, however, these are using very low number counts and should be treated carefully.
